EGGs are adhesive!

### 2025-07-30 jlamp/jlamp.tex
minima

\begin{abstract}
The use of rewriting-based visual formalisms is on the rise. 
%
In the formal methods community, this is due also to the introduction of adhesive
categories, where most properties of classical approaches to graph transformation, 
such as those on parallelism and confluence, can be rephrased and proved in a general and 
uniform way.
%
E-graphs (EGGs) are a formalism for program optimisation 
via an efficient implementation of equality saturation. 
In short, EGGs can be  defined as (acyclic) term graphs with an additional notion of 
equivalence on nodes that is closed under the operators of the signature.
Instead of replacing the components of a program, the optimisation step 
is performed by adding new components and linking them to 
the existing ones via an equivalence relation, until an optimal program is reached.
%
This work describes EGGs via adhesive categories. 
Besides the benefits in itself of a formal presentation, which renders the 
properties of the data structure precise, the description of the addition of equivalent 
program components using standard graph transformation tools offers the advantages 
of the adhesive framework in modelling, for example, concurrent updates.
%
%Besides 
%the benefits \emph{per se} of a formal presentation, making precise the properties of 
%the data structure, describing the addition of equivalent program components 
%via the standard tools of graph transformation offers the advantages given by the 
%adhesive framework in modelling e.g. concurrent updates.
\end{abstract}

@@

\section{Introduction}
The introduction of \emph{adhesive categories} marked a watershed moment for the algebraic approaches 
to the rewriting of graph-like structures~\cite{lack2005adhesive,ehrig2006fundamentals}.
%
Until then, key results of the approaches on e.g. parallelism and confluence had to be proven 
over and over again for each different formalism at hand, %let it be either xxx or xx, 
despite the obvious similarity of the procedure.
%
Adhesive categories provides such a disparate set of formalisms with a common abstract framework 
where many of these general results can be recast and uniformly proved once and for all.

\vspace{.1cm}
\noindent
\begin{minipage}[l]{.78\linewidth}
		\setlength{\parindent}{1.5em}
		In short, following the double-pushout (DPO) approach
to graph transformation~\cite{CorradiniMREHL97,ehrig2006fundamentals}, 
%~\cite{ehrig1999handbook,ehrig2006fundamentals,corradini1997algebraic}, 
a rule is given by two arrows $l: K \to L$ and $r: K \rightarrow R$
and its application requires a match $m: L \to G$: the rewriting step from $G$
to $H$ is then given by the diagram aside, whose halves are pushouts.
  \end{minipage}%
    \hfill 
  \begin{minipage}[r]{.20\linewidth }
    \xymatrix@C=.5cm@R=.5cm{
      L \ar[d]_{m}
      & K \ar[r]^r \ar[l]_{l} \ar[d]_{k} & R \ar [d]^h \\
      G & C \ar[r]_{g} \ar[l]^{f}                    & H
    }
%    \xymatrix@C=.5cm@R=.5cm{
%      L \ar[d]_{m} \ar@{}[dr]|{(1)}
%      & I \ar[r]^r \ar@{>->}[l]_{l} \ar[d] \ar@{}[dr]|{(2)} & R \ar [d] \\
%      G & C \ar@{->}[r] \ar@{>->}[l]                    & H\\
 %     & J \ar@{->}[ul] \ar@{->}[u]|{k} \ar@{->}[ur]
 %   }
  \end{minipage}

Thus, $L$ and $R$ are the left- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet that are not affected by the rule itself.
%
Should a category be adhesive, and the arrows of the rules monomorphisms, the presence of a match ensures that
if the pushout complement $C \to G$ exists then it is unique, hence a rewriting step can be deterministically performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can be recovered~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the theory is still under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness,
% and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.

@@

EGGs can be concisely defined as term graphs equipped with a notion of equivalence on nodes
that is closed under the operators of a signature~\cite[Section~4.2]{DetlefsNS05}.
%(i.e., graphs whose components are labelled with the operators of a signature)
In the presentation of term graphs via string diagrams~\cite{CastelnovoGM24}, EGGs are (hyper)trees 
whose edges are labelled by operators and with
the possibility of sharing subtrees, with an additional equivalence relation $\equiv$ on nodes that 
is closed under composition. In plain words and using a toy example:
if $a$ and $b$ are two constants such that $a \equiv b$, then $f(a) \equiv f(b)$ for any unary operator $f$.

@@

Building on the criterion developed in~\cite{CastelnovoGM24}, this work proves that both hypergraphs with equivalence
and EGGs form an 
$\mathcal{M}$-adhesive category for a suitable choice of $\mathcal{M}$.
The advantages from this characterisation are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the optimisation steps via the DPO approach
offers the tools for modelling their parallel and concurrent execution
and for proving their confluence and termination.

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for proving the $\mathcal{M}$-adhesivity
of hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} of
their variants where equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the machinery at work, showing how the optimisation steps
can be rephrased as the application of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, the proofs appear in the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

We open this background section by fixing some notation. Given a category $\X$ we do not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  Given an object $X$, we  denote by $?_X$ the unique arrow from an initial object into $X$ and by $!_X$ that  unique arrow from $X$ into a terminal one. We will also use the notation $e\colon X\eto Y$ to denote that an arrow $e\colon X\to Y$ is a regular epi.

@@

\begin{defi}[\cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}]
	Let $\X$ be a category and $\mathcal{M}$ a subclass of
	$\mon(\X)$  including  all isos, closed under composition, decomposition,  and stable under pullbacks and pushouts.  The category  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
	\begin{enumerate}
		\item it has \emph{$\mathcal{M}$-pullbacks}, i.e.~pullbacks along arrows of $\mathcal{M}$;
		\item it has \emph{$\mathcal{M}$-pushouts}, i.e.~pushouts along arrows of $\mathcal{M}$;
		\item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
	\end{enumerate}
	%
	A category $\X$ is said to be \emph{strictly $\mathcal{M}$-adhesive}
	if $\mathcal{M}$-pushouts are Van Kampen. We write $m\colon X \mto Y$ to denote that an arrow $m\colon X\to Y$ belongs to $\mathcal{M}$.
\end{defi}

@@

\begin{prop}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category, $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to \A$ one preserving $\mathcal{M}$-pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/\X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{25pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ that is stable under pullbacks and pushouts, contains all isos, and is closed under composition and decomposition.
	\end{enumerate} 
\end{prop}

@@

\begin{exa}
	\label{ex:adhesive}
	$\cat{Set}$ is adhesive, and, more generally, every topos is
	adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
	$\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
	where $E \rightrightarrows {V}$ is the two objects category with two
	morphisms $s,t \colon{E} \to {V}$. Similarly, various
	categories of hypergraphs can be shown to be adhesive, such as term
	graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, 
	i.e.~graphs without parallel edges, is
	$\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
	quasiadhesive.
\end{exa}

@@

Regular epis are particular well-behaved with respect to their kernel pairs.

@@

\noindent
\begin{minipage}[l]{.83\linewidth}
	\begin{rem}\label[remark]{rem:length} It is well known that $(\mathbb{N}), \{\delta_n\}_{n\in \mathbb{N}})$ is a coproduct for $\aleph_0$ copies of $1$, where $\delta_n\colon 1\to \mathbb{N}$ is the function picking the element $n$. Thus by the first point of \Cref{prop:fact} there is a canonical isomorphism $\Delta\colon \mathcal{N}\to 1^\star$. In particular, every set $X$ the arrow $!_\X\colon X\to 1$ induces a \emph{length function} $\lgh_{X}\colon X^\star \to \mathbb{N}$, which sends a word to its length, as in the diagram aside.
	\end{rem}
\end{minipage}\hfill \begin{minipage}[r]{.15\linewidth}\noindent
\xymatrix@R=10pt{1 \ar[r]^{\id{1}} \ar[d]^{u_n}& 1 \ar[d]^{v_n}\\ \mathbb{N} \ar@{.>}[r]^{\Delta}& 1^\star} \hspace{1pt}\\
\xymatrix@R=10pt@C=22pt{\mathbb{N} \ar[r]^{\Delta} \ar@{.>}[dr]_{\lgh_{X}}& 1^\star\\ & X^\star \ar[u]_{!^\star_X}}
\end{minipage}

@@

Note that by, \Cref{prop:fact}, $(-)^\star$ preserves pullbacks, while $\pro$ is continuous by definition, 
hence by \Cref{prop:com} and \Cref{cor:mono} we can deduce the following result.

@@

\Cref{prop:left,prop:com} allow us to deduce immediately the following.

@@

\begin{exa}\label[example]{exa_2}
We represent hypergraphs visually: dots denote nodes and boxes hyperedges. Should we be interested in their identity, we put a name near the corresponding dot or box. 
Sources and targets are represented by lines between dots and squares: the lines from the sources of a hyperedge comes from the left of the box, while the lines to the targets 
exit from the right of the box.
Let us look at the picture below. It represent a hypergraph $\mathcal{G}$ with sets $V_\mathcal{G} = \{v_1, v_2, v_3, v_4\}$ and $E_\mathcal{G} = \{h_1, h_2, h_3, h_4\}$ of nodes 
and edges, respectively, such that  $h_1, h_2$ have no source and $h_3, h_4$ a pair of nodes
	\begin{center}
        \begin{tikzpicture}
                \begin{pgfonlayer}{nodelayer}
                        \node[style=none] (lab1) at (-1.8, 0.8) {$h_1$};
                        \node[style=small box] (h1) at (-1.8, 0.3) {};
                        \node[style=small box] (2) at (-1.8, -0.5) {};
                        \node[style=none] (lab2) at (-1.8, -1.0) {$h_2$};
                        \node[style=none] (h1b) at (-1.8, 0.3) {};
                        \node[style=none] (th2) at (0, 0.3) {};
                        \node[style=none] (bh2) at (0, -0.3) {};
                        \node[style=none] (h2b) at (0, 0) {};
                        \node[style=none] (th3) at (1.8, 0){};
                        \node[style=none] (bh3) at (1.8, -0.6){};
                        \node[style=none] (h3c) at (1.8, -0.3){};
                        \node[style=none] (2c) at (-1.8, -0.5){};
                        \node[style=none] (lab3) at (0, 0.8) {$h_3$};
                        \node[style=medium box] (h2) at (0, 0) {};
                        \node[style=none] (lab4) at (1.8, 0.5) {$h_4$};
                        \node[style=medium box] (h3) at (1.8, -0.3) {};
                        \node[style=none](labv2) at (-0.9, -0.9) {$v_2$};
                        \node[style=node] (v2) at (-0.9, -0.5) {};
                        \node[style=none] (labv1) at (-0.9, 0.7) {$v_1$};
                        \node[style=node] (v1) at (-0.9, 0.3) {};
                        \node[style=none] (labv3) at (0.9, 0.4) {$v_3$};
                        \node[style=node] (v3) at (0.9, 0) {};
                        \node[style=none] (labv4) at (2.7, 0.1) {$v_4$};
                        \node[style=node] (v4) at (2.7, -0.3) {};
                \end{pgfonlayer}
                \begin{pgfonlayer}{edgelayer}
                        \draw (h1b.center) to (th2.center);
                        \draw (2c.center) to (v2);
                        \draw[in=180, out=30] (v2) to (bh2.center);
                        \draw (h2b.center) to (th3.center);
                        \draw (v4) to (h3c.center);
                        %\draw (h3c) to (v4);
                        \draw[in=180, out=-30](v2) to (bh3);
                \end{pgfonlayer}
        \end{tikzpicture}
	\end{center}
\end{exa}

>>>>

\begin{abstract}
The use of rewriting-based visual formalisms is on the rise. 
%
In the formal methods community, this is due also to the introduction of adhesive
categories, where most properties of classical approaches to graph transformation, 
such as those on parallelism and confluence, can be rephrased and proved in a general and 
uniform way.
%
E-graphs (EGGs) are a formalism for program optimisation 
via an efficient implementation of equality saturation. 
In short, EGGs can be  defined as (acyclic) term graphs with an additional notion of 
equivalence on nodes that is closed under the operators of the signature.
Instead of replacing the components of a program, the optimisation step 
is performed by adding new components and linking them to 
the existing ones via an equivalence relation, until an optimal program is reached.
%
This work describes EGGs via adhesive categories. 
Besides the usual benefits of a formal presentation, which makes the 
properties of a data structure precise, describing the addition of equivalent 
program components using standard graph transformation tools offers the advantages 
of the adhesive framework in modelling, for example, concurrent updates.
%
%Besides 
%the benefits \emph{per se} of a formal presentation, making precise the properties of 
%the data structure, describing the addition of equivalent program components 
%via the standard tools of graph transformation offers the advantages given by the 
%adhesive framework in modelling e.g. concurrent updates.
\end{abstract}

@@

\section{Introduction}
The introduction of \emph{adhesive categories} marked a watershed moment for the algebraic approaches 
to the rewriting of graph-like structures~\cite{lack2005adhesive,ehrig2006fundamentals}.
%
Until then, key results of the approaches on e.g. parallelism and confluence had to be proven 
over and over again for each different formalism at hand, %let it be either xxx or xx, 
despite the obvious similarity of the procedure.
%
Adhesive categories provides such a disparate set of formalisms with a common abstract framework 
where many general results can be recast and uniformly proved once and for all.

%\vspace{.1cm}
\noindent
\begin{minipage}[l]{.78\linewidth}
		\setlength{\parindent}{1.5em}
		In short, following the double-pushout (DPO) approach
to graph transformation~\cite{CorradiniMREHL97,ehrig2006fundamentals}, 
%~\cite{ehrig1999handbook,ehrig2006fundamentals,corradini1997algebraic}, 
a rule is given by two arrows $l: K \to L$ and $r: K \rightarrow R$
and its application requires a match $m: L \to G$: the rewriting step from $G$
to $H$ is then given by the diagram aside, whose halves are pushouts.
  \end{minipage}%
    \hfill 
  \begin{minipage}[r]{.20\linewidth }
    \xymatrix@C=.5cm@R=.5cm{
      L \ar[d]_{m}
      & K \ar[r]^r \ar[l]_{l} \ar[d]_{k} & R \ar [d]^h \\
      G & C \ar[r]_{g} \ar[l]^{f}                    & H
    }
%    \xymatrix@C=.5cm@R=.5cm{
%      L \ar[d]_{m} \ar@{}[dr]|{(1)}
%      & I \ar[r]^r \ar@{>->}[l]_{l} \ar[d] \ar@{}[dr]|{(2)} & R \ar [d] \\
%      G & C \ar@{->}[r] \ar@{>->}[l]                    & H\\
 %     & J \ar@{->}[ul] \ar@{->}[u]|{k} \ar@{->}[ur]
 %   }
  \end{minipage}

Thus, $L$ and $R$ are the left- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet that are not affected by the rule itself.
%
Should a category be adhesive, and the arrows of the rules monomorphisms, the presence of a match ensures that
if the pushout complement $C \to G$ exists then it is unique, hence a rewriting step can be deterministically performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can be recovered~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the theory is still under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness,
% and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, and for the $\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.

@@

EGGs can be concisely defined as (acyclic) term graphs equipped with an equivalence on nodes
that is closed under the operators of a signature~\cite[Section~4.2]{DetlefsNS05}.
%(i.e., graphs whose components are labelled with the operators of a signature)
In the presentation of term graphs via string diagrams~\cite{CastelnovoGM24}, EGGs are (hyper)trees 
whose edges are labelled by operators and with
possibly sharing subtrees, with an equivalence relation $\equiv$ on nodes that 
is closed under composition. In plain words and with a toy example:
if $a$ and $b$ are two constants such that $a \equiv b$, then $f(a) \equiv f(b)$ for any unary operator $f$.

@@

Building on the criterion developed in~\cite{CastelnovoGM24}, this work proves that both hypergraphs with equivalence
and EGGs form an 
$\mathcal{M}$-adhesive category for a suitable choice of $\mathcal{M}$.
The advantages from this characterisation are two-fold. On the one side, 
the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the optimisation steps via the DPO approach
offers the tools for modelling their parallel and concurrent execution
and for proving their confluence and termination.

\emph{Synopsis.}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for proving the $\mathcal{M}$-adhesivity
of hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} of
their variants where equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the machinery at work, showing how the optimisation steps
can be rephrased as the application of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, the proofs appear in the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

\emph{Differenza con la versione da conferenza}

@@

We open the background section by fixing notation. Given a category $\X$ we do not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  
Given an object $X$, we  denote by $?_X$ the unique arrow from an initial object into $X$ and by $!_X$ that  unique arrow from $X$ into a terminal one. 
The notation $e\colon X\eto Y$ denotes that an arrow $e\colon X\to Y$ is a regular epi.

@@

\begin{defi}[\cite{azzi2019essence,ehrig2012,ehrig2014adhesive,heindel2009category,lack2005adhesive}]
	Let $\X$ be a category and $\mathcal{M}$ a subclass of
	$\mon(\X)$  including  all isos, closed under composition, decomposition,  and stable under pullbacks and pushouts.  The category  $\X$ is said to be \emph{$\mathcal{M}$-adhesive} if
	\begin{enumerate}
		\item it has \emph{$\mathcal{M}$-pullbacks}, i.e.~pullbacks along arrows of $\mathcal{M}$;
		\item it has \emph{$\mathcal{M}$-pushouts}, i.e.~pushouts along arrows of $\mathcal{M}$;
		\item  $\mathcal{M}$-pushouts are $\mathcal{M}$-Van Kampen squares.
	\end{enumerate}
	%
	A category $\X$ is said to be \emph{strictly $\mathcal{M}$-adhesive}
	if $\mathcal{M}$-pushouts are Van Kampen squares. We write $m\colon X \mto Y$ to denote that an arrow $m\colon X\to Y$ belongs to $\mathcal{M}$.
\end{defi}

@@

\begin{prop}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category, $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to \A$ 
		a functor preserving $\mathcal{M}$-pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/\X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in \mathcal{M}\} \hspace{25pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		\[\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}\];
		\item if $\Y$ is a full subcategory of $\X$ closed under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ that is stable under pullbacks and pushouts, contains all isos, and is closed under composition and decomposition.
	\end{enumerate} 
\end{prop}

@@

\begin{exa}
	\label{ex:adhesive}
	$\cat{Set}$ is adhesive, and in fact every topos is
	adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
	$\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
	where $E \rightrightarrows {V}$ is the two objects category with two
	morphisms $s,t \colon{E} \to {V}$. Similarly, various
	categories of hypergraphs can be shown to be adhesive, such as term
	graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, 
	i.e.~graphs without parallel edges, is
	$\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
	quasiadhesive.
\end{exa}

@@

Regular epis are particularly well-behaved with respect to their kernel pairs.

@@

\noindent
\begin{minipage}[l]{.83\linewidth}
	\begin{rem}\label[remark]{rem:length} It is well known that $(\mathbb{N}, \{\delta_n\}_{n\in \mathbb{N}})$ is a coproduct for $\aleph_0$ copies of $1$, where $\delta_n\colon 1\to \mathbb{N}$ is the function picking the element $n$. Thus by the first point of \Cref{prop:fact} there is a canonical isomorphism $\Delta\colon \mathcal{N}\to 1^\star$. 
	In particular, for every set $X$ the arrow $!_\X\colon X\to 1$ induces a \emph{length function} $\lgh_{X}\colon X^\star \to \mathbb{N}$, which sends a word to its length, as in the diagram aside.
	\end{rem}
\end{minipage}\hfill \begin{minipage}[r]{.15\linewidth}\noindent
\xymatrix@R=10pt{1 \ar[r]^{\id{1}} \ar[d]^{u_n}& 1 \ar[d]^{v_n}\\ \mathbb{N} \ar@{.>}[r]^{\Delta}& 1^\star} \hspace{1pt}\\
\xymatrix@R=10pt@C=22pt{\mathbb{N} \ar[r]^{\Delta} \ar@{.>}[dr]_{\lgh_{X}}& 1^\star\\ & X^\star \ar[u]_{!^\star_X}}
\end{minipage}

@@

Note that by \Cref{prop:fact}, $(-)^\star$ preserves pullbacks, while $\pro$ is continuous by definition, 
hence by \Cref{prop:com} and \Cref{cor:mono} we can deduce the following result.

@@

\Cref{prop:left} and \Cref{prop:com} allow us to deduce immediately the following.

@@

\begin{exa}\label[example]{exa_2}
We represent hypergraphs visually: dots denote nodes and boxes hyperedges. Should we be interested in their identity, we put a name near the corresponding dot or box. 
Sources and targets are represented by lines between dots and squares: the lines from the sources of a hyperedge comes from the left of the box, while the lines to the targets 
exit from the right of the box.
Let us look at the picture below. It represent a hypergraph $\mathcal{G}$ with sets $V_\mathcal{G} = \{v_1, v_2, v_3, v_4\}$ and $E_\mathcal{G} = \{h_1, h_2, h_3, h_4\}$ of nodes 
and edges, respectively, such that  $h_1, h_2$ have no source and $h_3, h_4$ have a pair of nodes as soruce
	\begin{center}
        \begin{tikzpicture}
                \begin{pgfonlayer}{nodelayer}
                        \node[style=none] (lab1) at (-1.8, 0.8) {$h_1$};
                        \node[style=small box] (h1) at (-1.8, 0.3) {};
                        \node[style=small box] (2) at (-1.8, -0.5) {};
                        \node[style=none] (lab2) at (-1.8, -1.0) {$h_2$};
                        \node[style=none] (h1b) at (-1.8, 0.3) {};
                        \node[style=none] (th2) at (0, 0.3) {};
                        \node[style=none] (bh2) at (0, -0.3) {};
                        \node[style=none] (h2b) at (0, 0) {};
                        \node[style=none] (th3) at (1.8, 0){};
                        \node[style=none] (bh3) at (1.8, -0.6){};
                        \node[style=none] (h3c) at (1.8, -0.3){};
                        \node[style=none] (2c) at (-1.8, -0.5){};
                        \node[style=none] (lab3) at (0, 0.8) {$h_3$};
                        \node[style=medium box] (h2) at (0, 0) {};
                        \node[style=none] (lab4) at (1.8, 0.5) {$h_4$};
                        \node[style=medium box] (h3) at (1.8, -0.3) {};
                        \node[style=none](labv2) at (-0.9, -0.9) {$v_2$};
                        \node[style=node] (v2) at (-0.9, -0.5) {};
                        \node[style=none] (labv1) at (-0.9, 0.7) {$v_1$};
                        \node[style=node] (v1) at (-0.9, 0.3) {};
                        \node[style=none] (labv3) at (0.9, 0.4) {$v_3$};
                        \node[style=node] (v3) at (0.9, 0) {};
                        \node[style=none] (labv4) at (2.7, 0.1) {$v_4$};
                        \node[style=node] (v4) at (2.7, -0.3) {};
                \end{pgfonlayer}
                \begin{pgfonlayer}{edgelayer}
                        \draw (h1b.center) to (th2.center);
                        \draw (2c.center) to (v2);
                        \draw[in=180, out=30] (v2) to (bh2.center);
                        \draw (h2b.center) to (th3.center);
                        \draw (v4) to (h3c.center);
                        %\draw (h3c) to (v4);
                        \draw[in=180, out=-30](v2) to (bh3);
                \end{pgfonlayer}
        \end{tikzpicture}
	\end{center}
\end{exa}
